%
% %CopyrightBegin%
%
% Copyright Ericsson AB 2017. All Rights Reserved.
%
% Licensed under the Apache License, Version 2.0 (the "License");
% you may not use this file except in compliance with the License.
% You may obtain a copy of the License at
%
%     http://www.apache.org/licenses/LICENSE-2.0
%
% Unless required by applicable law or agreed to in writing, software
% distributed under the License is distributed on an "AS IS" BASIS,
% WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
% See the License for the specific language governing permissions and
% limitations under the License.
%
% %CopyrightEnd%
%

\chapter*{``Release notes'' for \Draft{0.7}}

This draft has the following important differences from \Draft{0.6}:

\begin{itemize}
\item Cleaned up meaning of \Erlang.
\item Removed \StdErlang\ lexical rules from lexical summary. (Used
Std test in ALL \Std\ rules)
\item Switch expression of case fixed.
\item LineTerminator is only LF in \OldErlang.  Warn about using
whitespace after \TXT{\$}.
\item Fix recognizer BIFs in \OldErlang\ to be only guard
recognizers.
\item Re-did macro definition and application syntax.  Hopefully
correct.
\item All numeric literals (Integer, Decimal, ExplicitRadix and Float)
are now unsigned in the lexical structure.
\item Made the logical operators have the same priorities as the
additive and multiplicative operators, as they should.
\item Atomic literals now defined in the expression chapter as this is
where values are explained.
\end{itemize}

\chapter*{``Release notes'' for \Draft{0.6}}

This draft has the following important differences from \Draft{0.5}:

\begin{itemize}
\item There is now an index (laboriously compiled).
\item There is a neat summary of all \Erlang\ expressions in \S\ref{chapter:summary-exprs}.
\item Function application is now described.
\item Records are now described.
\item Ports are now fully (?) described.
\item Dynamic code loading is now fully (?) described.
\item The external term format is now described.
\item The portable (?) hash function is now described.
\item Parse trees are now described.
\item Added \T{try \Z{E} end} to be almost equivalent with
\T{catch \Z{E}}.
\item The description of atoms has changed (\S\ref{section:atoms}).
\item All BIFs of \OldErlang\ are described (\S\ref{chapter:bifs}), except two:
\T{set_node/2} and \T{set_node/3}.
\item Exact equality and arithmetic equality (formerly called coerced equality)
are now well-defined, I think (\S\ref{section:equality}).
\item The section about Unicode escapes \ifStd (\S\ref{section:unicode-escapes}) \fi
is now written.
\item Lowered minimum \I{maxint} to $2^{59}-1$ (it was $2^{63}-1$)
so that on a 64-bit machine
an implementation without bignums can store all numbers as scalars with four
bits for a tag etc.
%\item The exit cause \T{badarith} has disappeared and been replaced by \T{badarg}.
\item Operators \T{//} and \T{mod} have been added (computing the same as
\T{div} and \T{rem} except that they round towards negative infinity rather
than zero).
\item Extent of function calls and last call optimization has been defined
(\S\ref{section:extent-function-calls}).
\item Dropped type and rule declarations completely from the specification.
\item Scheduling and waiting in \T{receive} expressions is described more precisely.
\end{itemize}

I am considering:
\begin{itemize}
\item Moving \S\ref{section:coding-pattern-matching} to an appendix.
\end{itemize}

Note:
\begin{itemize}
\item In the version for \StdErlang, I have assumed that the proposal
about new BIFs will be accepted, hence there are quite a lot of
references to such BIFs.  However, they are not described yet, hence
there are also quite a lot of unresolved cross-references in the
\StdErlang\ version.  If the BIF proposal is not accepted, then
I will change to use the current BIFs also in that version, of course.
\end{itemize}

The following is what remains:
\begin{itemize}
\item Checking release notes for R4.
\item The BIFs \T{set_node/2} (\S\ref{section:setnode2}) and
\T{set_node/3} (\S\ref{section:setnode3}).
\item Adding appendix about EPMD and \Erlang\ to \Erlang\ node communication.
\item If remaining proposals are accepted, incorporating them (the new BIFs is a hog,
the rest is fairly easy).
\item Checking that monitoring of nodes is described properly.
\end{itemize}

PLEASE, DRAFT READERS: THERE ARE PLACES IN THE TEXT WHERE I REQUEST
INFORMATION, ESPECIALLY ABOUT VARIOUS PARAMETERS OF ERLANG 4.7.  IF
YOU KNOW THESE THINGS, EMAIL ME!

\chapter*{``Release notes'' for \Draft{0.5}}

This draft has the following important differences from \Draft{0.4}:

\begin{itemize}
\item It has been added to the introduction how the specification
relates to the \Erlang\ language and the implementations \OldErlang\
and \NewErlang.
\item Some notions have been added to the glossary: expression,
operator, macro and term.
\item Macro \T{?VERSION} added.  Comments?
\item The description of equality tests is not quite accurate, will rewrite.
\item The evaluation order has been updated to be strictly left-to-right.
\item The section about guards (\S\ref{section:guards}) has been rewritten on
considerably simpler form.
\item The directionality when relating input and output environments has been made more clear.
\item It should now be even clearer that clauses are tried left to right and that
in \T{receive} expressions, all clauses are tried for a term in the message queue before
the next term is tried.
\item The terminology for abrupt completion has been made more uniform.
\item New chapter (\S\ref{chapter:arithmetics}) about arithmetics (LIA-1 based).
\item New (but not yet complete) chapter (\S\ref{chapter:more-about-ports}) about ports.
\item The format for BIF descriptions has evolved and many BIFs described.
\item The section about scheduling (\S\ref{section:scheduling}) has been improved.
\item New (short) sections about lifetime of terms (\S\ref{section:life-time}) and about
memory management (\S\ref{section:memory-management}).
\end{itemize}

Some things that are planned to be improved (and thus do not need to
be commented upon in the \Draft{0.5} if you agree):
\begin{itemize}
\item I will try to reduce the number of forward references further.
\end{itemize}

Known unclear issues:
\begin{itemize}
\item How should shadowing of BIFs work for those that are guard BIFs?
\end{itemize}

I have not gone through all comments from Torbj\"{o}rn Keisu.  They
should be incorporated in release 0.5.1.

\chapter*{``Release notes'' for \Draft{0.4}}

This draft has the following important differences from \Draft{0.3.1}:

\begin{itemize}
\item `\TXT{.}'\ and `\TXT{?}' are now listed as separators.
\item Full stops are detected properly (before white space and comments have been thrown
away).
\item The structure of a module is now described as consisting of
\begin{itemize}
\item First a module declaration \T{-module(\Z{ModuleName})}.
\item Then any \emph{header forms}, i.e., \T{export}, \T{import} and \T{compile}
attributes, together with \T{type} and \T{rule} declarations.
\item Then the \emph{program forms}, i.e., function declarations mixed with
\T{file} and wild attributes, record declarations and macro definitions.
\end{itemize}
\item The compilation of \Erlang\ programs is now described in detail in
\S\ref{chapter:compilation}.
\item A \emph{list} is now either \T{[]} or a cons in which the right part is a list.
\item The description of coercion has been improved (\S\ref{section:coercion}).
\end{itemize}

I have taken (most of) the written comments on 0.3 and/or 0.3.1 by
HÂkan and Per Mildner into account.

Some things that are planned to be improved (and thus do not need to
be commented upon in the \Draft{0.4} if you agree):
\begin{itemize}
\item Express better the directionality when relating input and output environments
of expressions.
\item Express arithmetics in terms of LIA-1.
\item Change the writing so that becomes clear that clauses (in \T{case} etc) are
tried left to right.
\item Make the writing for \T{receive} expressions \emph{even} clearer concerning
the fact that for each message, each clause is tried before the next message is
considered.
\item Use a more consistent terminology for abrupt completion and its \\ causes.
\item The BIF Chapter (\S\ref{chapter:bifs}) is both incomplete and requiring changes.
\item Assuming that the revised proposal about evaluation order is accepted,
the evaluation order will be changed to strict left-to-right everywere.
\end{itemize}

\chapter*{``Release notes'' for \Draft{0.3.1}}

This draft has the following important differences from \Draft{0.3}:

\begin{itemize}
\item Various typos have been fixed (in particular the ones in \S\ref{section:match-expr}
and \S\ref{section:send-expr}
about match and send expressions and in \S\ref{section:moduleinfo1} about \T{module_info/1}
for \T{imports}).
\item The operator \T{and} is no longer also listed as a keyword
(\S\ref{section:keywords}).
\item The proposed new escape sequence \verb|\s| (for space) is described
(\S\ref{section:escapes}).
\item The blurb about process communcation has been removed from the
introduction of binaries (\S\ref{section:binaries}).
\item Records are now in the term order (between tuples and lists)
described in \S\ref{section:term-order}.
\item The \T{unregistered_name} error handler has been removed
(\S\ref{section:send-expr}), on request from Klacke and Tony.
\item The description of the list difference operator \T{--} has been
fixed so now it may be correct (\S\ref{section:list-subtraction}).
\item The description of \T{receive} expressions has been somewhat adjusted
to accord better with reality and with the intentions (\S\ref{section:receive-expr}).
\item The descriptions of how the clauses in \T{if}, \T{case}, \T{receive}, \T{try}
and \T{fun}
expressions are checked have been improved to increase clarity
(\S\ref{section:if-expr}, \S\ref{section:case-expr}, \S\ref{section:receive-expr},
\ifStd \S\ref{section:try-expr}, \fi \S\ref{section:fun-exprs}).
\item The description of \T{receive} expressions has been significantly
improved and may be correct (\S\ref{section:fun-exprs}).
\item The missing Section~\ref{section:guards} about guards is no longer missing.
\item \S\ref{chapter:processes} has been almost completely rewritten and
uses a new concept of \emph{signals} that comprises messages and exit signals,
as well as link/unlink requests, etc.  (This is actually how it is implemented
in the `5.0' system and the presentation becomes more precise.)
\item The priority \T{high} has been added (\S\ref{chapter:processes}).
\end{itemize}

I have taken the written comments on 0.3 by Klacke and Tony into account.
The comments by Per Mildner remain (I forgot to bring them to Stockholm).
